### 2025-08-06 cognitive-bias.qmd
20250806 - tweaks and regression to the mean

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called [*heuristics*](#sec-heuristics).
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
For instance, [heuristics](#sec-heuristics) can lead us to identify patterns out of randomness.
[*Cognitive biases*](#sec-cognitiveBiases) are systematic errors in thinking.
[*Fallacies*](#sec-fallacies) are forms of flawed reasoning (invoked as an argument).
Fallacies are a characteristic of an argument, whereas cognitive biases are characteristics of the reasoning agent.
Thus, [fallacies](#sec-fallacies) can result from [heuristics](#sec-heuristics) and [cognitive biases](#sec-cognitiveBiases).

@@

Applied to fantasy football, a manager is trying to predict how many fantasy points a top Running Back may score.
The player scored 300 fantasy points last season, but the team added a stronger backup Running Back and changed the Offensive Coordinator to be a more pass-heavy offense.
The manager may use 300 fantasy points as an anchor (based on the player's performance last season), and may adjust downward 15 points to account for the offseason changes.
However, it is possible that this downward adjustment is insufficient to account not only for the offseasons changes but also for potential [regression effects](#sec-fallaciesRegression).
[Regression effects](#sec-fallaciesRegression) are discussed further in @sec-fallaciesRegression.

@@

1. *overestimation* of one's actual performance
1. *overplacement* of one's performance relative to others
1. *overprecision* in one's beliefs/jugments/predictions

@@

*Hindsight bias* is the tendency to perceive that past events were more predictable than they were (i.e., the "I-knew-it-all-along" effect).
Everything seems to make sense in hindsight [@Kahneman2011].
People tend to remember the succeess of their predictions and forget the failures of their predictions.
For instance, if a third-string Quarterback has a breakout game, a fantasy manager may claim that they "knew it all along" that the player was going to breakout, despite not having picked up the player.
That same manager may forget the many other predictions they had that did not come true.

@@

*Self-serving bias* is the tendency to perceive oneself in an overly positive manner.
Doing so may function to bolster people's self-esteem.
One way that self-serving bias can commonly manifest is the tendency people to attribute successes to internal factors—such as one's character, ability, effort, or actions—but to attribute failures to external factors beyond one's control—such as others, unfairness, bad luck, etc.
That is, self-serving bias involves claiming more reponsibility for one's success than for one's failure [@Miller2013c].
For instance, a manager who wins the league may attribute their success to their fantasy football skill and smart decisions (rather than luck).
However, if they get last place, they may attribute their failure not to themselves but instead to factors outside of their control, such as bad luck, player injuries, and poor performance by players due to circumstances out of their control (e.g., bad weather conditions, coaching decisions).

@@

*Loss aversion bias* is the tendency to avoid losses rather than acquiring equivalent gains.
That is, loss aversion (similar to punishment sensitivity, as opposed to reward sensitivity) is the tendency to weigh losses (or punishment) more heavily than gains (or rewards).
Thus, the target state matters less than the *change* (i.e., gains and/or losses) from one's reference point [e.g., their current roster; @Kahneman2011].
Loss aversion is a key aspect of prospect theory and suggests that individuals tend to be biased in favor of their reference situation and toward making smaller rather than larger changes [@Kahneman2011].
Loss aversion explains why people tend to favor the favoring the status quo; the negative consequences of a change often loom larger than the positive consequences of a change [@Kahneman2011].
Aversion to losses and potential threats may be related to humans' survival instinct [@Kahneman2011].
With par as a reference point, this may explain why professional golfers tend to putt more accurately for par than for birdie, even for putts of equivalent distance [@Kahneman2011].

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
The, a "low base-rate" event is an event with a low probability of occurring (i.e., an unlikely event).
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
This is especially true if the unlikely event attracts attention, because of the [confirmatory bias](#sec-cognitiveBiasesConfirmation) of memory [@Kahneman2011].

@@

The regression fallacy is the failure to account for the fact that things tend to naturally fluctuate around their mean and that, after an extreme fluctuation, subsequent scores tend to regress (or reverse) to the mean.
Regression to the mean occurs because performance—both success and failure—is influenced by talent and by luck [@Kahneman2011].
And, unlike talent (or one's level one the construct of interest), luck is fickle.
Regression to the mean can explain why those with depression tend to improve over time without doing anything, and can explain why highly intelligent women tend to marry men who are less intelligent than they are [@Kahneman2011].
Despite peoples' tendency to ascribe such changes to causal phenomena, regression-related changes do not need a causal explanation—they are a mathematically inevitable consequence that arises from the fact that an outcome is influenced, to some degree, both by one's level on the construct and by error [@Kahneman2011].

>>>>

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called [*heuristics*](#sec-heuristics).
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
For instance, [heuristics](#sec-heuristics) can lead us to identify patterns out of randomness.
[*Cognitive biases*](#sec-cognitiveBiases) are systematic errors in thinking.
[Cognitive biases](#sec-cognitiveBiases) can result from [heuristics](#sec-heuristics).
[*Fallacies*](#sec-fallacies) are forms of flawed reasoning (invoked as an argument).
[Fallacies](#sec-fallacies) are a characteristic of an argument, whereas [cognitive biases](#sec-cognitiveBiases) are characteristics of the reasoning agent.
Thus, [fallacies](#sec-fallacies) can result from [heuristics](#sec-heuristics) and [cognitive biases](#sec-cognitiveBiases).

One might wonder why [fallacies](#sec-fallacies) and [cognitive biases](#sec-cognitiveBiases) exist, given that errors in thinking and reasoning would seem to be disadvantageous.
However, [fallacies](#sec-fallacies) and [cognitive biases](#sec-cognitiveBiases) reflect the tradeoffs between speed and accuracy in thinking and, in some cases, may also serve to help people feel better about themselves.

@@

Applied to fantasy football, a manager is trying to predict how many fantasy points a top Running Back may score.
The player scored 300 fantasy points last season, but the team added a stronger backup Running Back and changed the Offensive Coordinator to be a more pass-heavy offense.
The manager may use 300 fantasy points as an anchor (based on the player's performance last season), and may adjust downward 15 points to account for the offseason changes.
However, it is possible that this downward adjustment is insufficient to account not only for the offseasons changes but also for potential [regression effects](#sec-fallaciesRegression).
[Regression effects](#sec-fallaciesRegression) are discussed further in @sec-fallaciesRegression.
As another example, the first number or players named in a negotiation can serve as an anchor.

@@

1. *overestimation* of one's actual performance
1. *overplacement* of one's performance relative to others
1. *overprecision* in one's beliefs/judgments/predictions

@@

*Hindsight bias* is the tendency to perceive that past events were more predictable than they were (i.e., the "I-knew-it-all-along" effect).
Everything seems to make sense in hindsight [@Kahneman2011].
People tend to remember the successes of their predictions and forget the failures of their predictions.
For instance, if a third-string Quarterback has a breakout game, a fantasy manager may claim that they "knew it all along" that the player was going to breakout, despite not having picked up the player.
That same manager may forget the many other predictions they had that did not come true.

@@

*Self-serving bias* is the tendency to perceive oneself in an overly positive manner.
Doing so may function to bolster people's self-esteem.
One way that self-serving bias can commonly manifest is the tendency people to attribute successes to internal factors—such as one's character, ability, effort, or actions—but to attribute failures to external factors beyond one's control—such as others, unfairness, bad luck, etc.
That is, self-serving bias involves claiming more responsibility for one's success than for one's failure [@Miller2013c].
For instance, a manager who wins the league may attribute their success to their fantasy football skill and smart decisions (rather than luck).
However, if they get last place, they may attribute their failure not to themselves but instead to factors outside of their control, such as bad luck, player injuries, and poor performance by players due to circumstances out of their control (e.g., bad weather conditions, coaching decisions).

@@

*Loss aversion bias* is the tendency to avoid losses rather than acquiring equivalent gains.
That is, loss aversion (similar to punishment sensitivity, as opposed to reward sensitivity) is the tendency to weigh losses (or punishment) more heavily than gains (or rewards).
Thus, the target state matters less than the *change* (i.e., gains and/or losses) from one's reference point [e.g., their current roster; @Kahneman2011].
Loss aversion is a key aspect of prospect theory and suggests that individuals tend to be biased in favor of their reference situation and toward making smaller rather than larger changes [@Kahneman2011].
Loss aversion explains why people tend to favor the status quo; the negative consequences of a change often loom larger than the positive consequences of a change [@Kahneman2011].
Aversion to losses and potential threats may be related to humans' survival instinct [@Kahneman2011].
With par as a reference point, this may explain why professional golfers tend to putt more accurately for par than for birdie, even for putts of equivalent distance [@Kahneman2011].

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Thus, a "low base-rate" event is an event with a low probability of occurring (i.e., an unlikely event).
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
This is especially true if the unlikely event attracts attention, because of the [confirmatory bias](#sec-cognitiveBiasesConfirmation) of memory [@Kahneman2011].

@@

The regression fallacy is the failure to account for the fact that things tend to naturally fluctuate around their mean and that, after an extreme fluctuation, subsequent scores tend to regress (or reverse) to the mean.
Regression to the mean is depicted in @fig-regressionToMean.
Regression to the mean occurs because performance—both success and failure—is influenced by talent and by luck [@Kahneman2011].
And, unlike talent (or one's level one the construct of interest), luck is fickle.
Regression to the mean can explain why those with depression tend to improve over time without doing anything, and can explain why highly intelligent women tend to marry men who are less intelligent than they are [@Kahneman2011].
Despite peoples' tendency to ascribe such changes to causal phenomena, regression-related changes do not need a causal explanation—they are a mathematically inevitable consequence that arises from the fact that an outcome is influenced, to some degree, both by one's level on the construct and by error [@Kahneman2011].

```{r}
#| label: fig-regressionToMean
#| fig-cap: "Regression to the Mean."
#| fig-alt: "Regression to the Mean."
#| code-fold: true

mydata <- data.frame(
  week = 1:7,
  fantasyPoints = c(25,7,12,5,11,15,3)
)

ggplot(
  data = mydata,
  mapping = aes(
    x = week,
    y = fantasyPoints
  )
) + 
  geom_point(
    size = 4
  ) +
  geom_line(
    linewidth = 1
  ) +
  geom_abline(
    intercept = 10,
    slope = 0,
    linetype = "dashed"
  ) +
  scale_x_continuous(
    breaks = 1:7
  ) +
  labs(
    y = "Fantasy Points",
    title = "Regression to the Mean"
  ) +
  annotate(
    "segment",
    x = 6.5,
    xend = 7,
    y = 17,
    yend = 10,
    color = "blue",
    linewidth = 1.5,
    alpha = 0.6,
    arrow = arrow()) +
  annotate(
    "text",
    x = 5.5,
    y = 18,
    label = "Average ability",
    hjust = 0, # left-justify
    size = 5) +
  theme_classic(base_size = 30) +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```
